### 2024-03-29 manuscript.Rmd
update analysis and results

The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] with exceptionally diverse morphology from slender (aspect ratio = ) to robust. Body proportion exhibit strong phylogenetic signal but in some instances spread out across the phylogeny (CITE).

In this study, we investigate the interspecific variation in the number of pre- and post-cloacal vertebrae to test the hypothesis that changes in developmental patterns underpin the evolution of different body proportions in Australian typhlopid snakes.

Specifically, we test two predictions:
(1) narrow-and-long species develop from postnatal ontogenetic changes where vertebral numbers are not significantly different between narrow-and-long species compared to wide-and-robust species; but are elongated.

(2) Narrow-and-long blindsnakes develop via changes in the embryonic and postnatal stages where narrow-and-long species have greater number of vertebrae than wide-and-robust species.

@@

## Phylogeny
We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in [@tiatragulAustralia2023]. The subset phylogeny represents 35/47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A. grypus_" and one eastern population of "_A. ligatus_" [@tiatragulAustralia2023]. This phylogeny was inferred using the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] based on 4930 maximum likelihood gene trees estimated using IQTREE [@minhIQTREE2020]. Divergence dating was estimated using MCMCTree from the PAML package [@yangPAML2007] based on 27 squamate reptile fossil calibrations [@tiatragulAustralia2023].

@@

We used the Fiji distribution of ImageJ2 [@schindelinFiji2012] to digitally visualise X-ray images and counted the number of vertebrae using Cell Counter module. Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023]. Additionally, we quantified intracolumnar length of each vertebrae for two representative blindsnake species (*A. grypus* and *A. ligatus*). Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra. We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R. Measurements were repeated three times to account for measurement errors.

@@

We selected soil bulk density, annual mean temperature, and aridity index as the most biologically relevant data to describe blindsnake niches. 
Soil bulk density measures the penetrability of soil, which is relevant for strictly fossorial reptiles [camaitiMacroecological2022]. Bulk density is calculated by taking the weight of dry soil divided by the total soil volume (g/cm<sup>3</sup>). Values > 1.6 g/cm<sup>3</sup> tend to restrict root growth, while loose soil have densities < 0.5 g/cm<sup>3</sup> [@mckenzieSoil2002]. Mean annual temperature serves as a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013]. Finally, aridity index reflects precipitation availability over atmospheric water demand and is categorised as follows: hyper arid, < 0.03; arid, 0.03 – 0.2; semi-arid, 0.2-0.5; dry sub-humid, 0.5 – 0.65; humid, >0.65 [@zomerVersion2022]. 
From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017], and aridity index from @zomerVersion2022 for each georeferenced record using the "extract" function in the R package `raster` v.3.6-3 [@hijmansRaster2022]. We then calculated the maximum soil bulk density, mean annual temperature, and mean aridity index for each species. Correlation analyses revealed that none of the background environmental variables were strongly correlated (Pearson's *r* < |0.75|). 
 -->
#### Statistical analyses

All statistical analyses were performed in R v.4.31 [@rcoreteamLanguage2023]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Browninan motion implemented in `geomorph` v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. 
We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum SVL from Tiatragul et al. (unpublished). Then, we test whether the total vertebrae number to total body length ratio correlates with body shape, defined as the ratio between midbody diameter divided by SVL, also from Tiatragul et al. (unpublished) data.

@@

We found no relationship between indicating maximum number of total vertebrae and maximum SVL (PGLS, slope = -0.08, y=225, R^2^=0.02, F=0.9, Z=0.40, P=0.361), indicating an absence of pleomerism in Australian blindsnakes.

Vertebrae ratio is negatively correlated with body width ratio (PGLS, slope = -21.4, y=1.4, R^2^=0.2, F=8.8,, Z=2.3, P=0.007), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-width-ratios)).

@@

<!-- Figure 2: Pleomerism and ratio correlations -->
```{r fig-vert-width-ratios, fig.cap= 'Relationship between mean vertebrae ratio and mean body width ratio among Australian blindsnakes. Vertebrae ratio is calculated by total vertebrae/total length, and body width ratio is midbody diameter/total body length.', out.width="165mm", out.height= "126mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/vert-width-ratios.jpg'))
```

>>>>

The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] with considerably diverse morphology from narrow-and-long (aspect ratio >= 100) to robust-and-wide (aspect ratio < 60).

In this study, we investigate the interspecific variation in the number of vertebrae to test the hypothesis that changes in developmental patterns underpin the evolution of different body shapes in Australian typhlopid snakes.

Specifically, we test two predictions:
(1) narrow-and-long species develop from postnatal ontogenetic changes where vertebral numbers are not significantly different between narrow-and-long species compared to wide-and-robust species; but are elongated. 
(2) Narrow-and-long blindsnakes develop via changes in the embryonic and postnatal stages where narrow-and-long species have greater number of vertebrae than wide-and-robust species.

@@

## Phylogeny

We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in [@tiatragulAustralia2023]. The subset phylogeny represents 35/47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A. grypus_" and one eastern population of "_A. ligatus_" [@tiatragulAustralia2023]. This phylogeny was inferred using the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] based on 4930 maximum likelihood gene trees estimated using IQTREE [@minhIQTREE2020]. Divergence dating was estimated using MCMCTree from the PAML package [@yangPAML2007] based on 27 squamate reptile fossil calibrations [@tiatragulAustralia2023].

@@

We used the Fiji distribution of ImageJ2 [@schindelinFiji2012] to digitally visualise X-ray images and counted the number of vertebrae using Cell Counter module. Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023]. Additionally, we quantified intracolumnar length of each vertebrae for three species representing narrow-and-wide phenotypes *A. grypus*, wide-and-robust phenotypes (*A. pinguis*), and intermediate phenotypes (*A. centralis*). Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra. We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R. Measurements were repeated three times to account for measurement errors.

@@

We selected soil bulk density, annual mean temperature, and aridity index as the most biologically relevant data to describe blindsnake niches. 
Soil bulk density measures the penetrability of soil, which is relevant for strictly fossorial reptiles [camaitiMacroecological2022]. Bulk density is calculated by taking the weight of dry soil divided by the total soil volume (g/cm<sup>3</sup>). Values > 1.6 g/cm<sup>3</sup> tend to restrict root growth, while loose soil have densities < 0.5 g/cm<sup>3</sup> [@mckenzieSoil2002]. Mean annual temperature serves as a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013]. Finally, aridity index reflects precipitation availability over atmospheric water demand and is categorised as follows: hyper arid, < 0.03; arid, 0.03 – 0.2; semi-arid, 0.2-0.5; dry sub-humid, 0.5 – 0.65; humid, >0.65 [@zomerVersion2022]. 
From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017], and aridity index from @zomerVersion2022 for each georeferenced record using the "extract" function in the R package `raster` v.3.6-3 [@hijmansRaster2022]. We then calculated the maximum soil bulk density, mean annual temperature, and mean aridity index for each species. Correlation analyses revealed that none of the background environmental variables were strongly correlated (Pearson's *r* < |0.75|). 
 -->

## Statistical analyses

All statistical analyses were performed in R v.4.31 [@rcoreteamLanguage2023]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Browninan motion implemented in `geomorph` v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. 
We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from Tiatragul et al. (unpublished). Then, we test whether the vertebrae to body length ratio is related to maximum aspect ratio, defined by total body length / body width, also from Tiatragul et al. (unpublished) data.

@@

We found a weakly positive but non statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, y=256, R^2^=0.02, F=0.72, Z=0.29, P=0.42), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig: fig-tbl-vert)).

However, we found that vertebrae ratio is postively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09,, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-width-ratios)).

@@

<!-- Fig. Vertebral counts vs max body length  -->
```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length for Australian typhlopids. Trend line represent PGLS regression for Australian species.', out.width="165mm", out.height= "126mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/tbl-vert.jpg'))
```

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= '(B) Relationship between mean vertebrae and aspect ratios among Australian blindsnake species. Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and wide', out.width="165mm", out.height= "126mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/vert-aspect.jpg'))
```
